Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2019-03-10 mpaper/paper.tex
Introduction to the theoretical results

@@

\subsection{Variational Inference} \label{vi_algs}

The typical way to optimise a quantity (the ELBO, in this case) involves
computing its gradient \cite{blei2017variational}. Unfortunately, some of the
terms in $\mathcal{L}$ are still left as expected values. \emph{Black box
  variational inference} (BBVI) \cite{DBLP:conf/aistats/RanganathGB14} suggests
a way to express the gradient as an expectation without having to take the
gradient of the posterior:
\[ \nabla_{\bm\nu} \mathcal{L} = \mathbb{E}[\nabla_{\bm\nu} \log \approximation
  (\log \pfull - \log \approximation)]. \]
The gradient of the ELBO then has a unbiased estimate
\[
  \nabla_{\bm\nu} \mathcal{L} \approx \frac{1}{S} \sum_{s=1}^S \nabla_{\bm\nu} \log \approximationS (\log \pfullS - \log \approximationS),
\]
computed by drawing $S$ Monte Carlo samples $(\mathbf{u}_s, \mathbf{r}_s) \sim
\approximation$.

The main theorem behind the BBVI trick is the Lebesgue's dominated convergence
theorem \cite{royden1988real}, which---under some conditions---allows one to
claim that:
\[ \lim_{n \to \infty} \int_E f_n\,d\mu = \int_E \lim_{n \to \infty}
  f_n\,d\mu. \]
We can use the same idea on parts of $\mathcal{L}$, derivatives of which cannot
be taken otherwise. In our case, we want a gradient instead of a limit, and the
integral represents an expected value.

%While the ELBO can now be computed without approximations, we ended up imposing
%two zero-variance restrictions, resulting in a less informative model. Moreover,
%we had to abandon the Bayesian treatment of $\bm\lambda$, making the model more
%vulnerable to overfitting.

@@

% TODO: reference in text
\begin{figure}
  \centering
  \begin{tikzpicture}
    \node [rectangle, draw, fill=lightgray] (perturbation) at (2, 0) {\textsc{Perturbation Lemma}};
    \node [rectangle, draw, fill=lightgray] (lebesgue) at (-1, -3) {\textsc{Dominated Convergence Theorem}};
    \node [rectangle, draw] (derivatives) at (-2, 0) {\textsc{Derivatives of PDFs}};
    \node [rectangle, draw] (measurability) at (-3, -2) {\textsc{Measurability}};
    \node [rectangle, draw] (boundedness) at (3, -2) {\textsc{Boundedness}};
    \node [rectangle, draw] (lemma1) at (-2, -1) {\textsc{Lemma \ref{lemma:bound1}}};
    \node [rectangle, draw] (lemma2) at (0, -1) {\textsc{Lemma \ref{lemma:bound2}}};
    \node [rectangle, draw] (lemma3) at (2, -1) {\textsc{Lemma \ref{lemma:bound3}}};
    \node [rectangle, draw] (lemma4) at (3, -3) {\textsc{Lemma \ref{lemma:integral_of_r}}};
    \node [rectangle, draw] (theorem) at (0, -2) {\textsc{Theorem \ref{thm:main}}};

    \draw [arrow] (perturbation) edge (lemma3);
    \draw [arrow] (lebesgue) edge (theorem);
    \draw [arrow] (derivatives) edge (lemma1);
    \draw [arrow] (derivatives) edge (lemma2);
    \draw [arrow] (derivatives) edge (lemma3);
    \draw [arrow] (measurability) edge (theorem);
    \draw [arrow] (boundedness) edge (theorem);
    \draw [arrow] (lemma1) edge (theorem);
    \draw [arrow] (lemma2) edge (theorem);
    \draw [arrow] (lemma3) edge (theorem);
    \draw [arrow] (lemma4) edge (theorem);
  \end{tikzpicture}
  \caption{A graphical representation of dependencies between our theoretical
    results. An arrow from $A$ to $B$ means that $A$ was used to prove $B$.
    Results from the literature are in gray.}
\end{figure}

MDP values are characterised by both a state and a reward function/vector. In
order to prove the next theorem, we think of the value function as $V :
\mathcal{S} \to \mathbb{R}^{|\mathcal{S}|} \to \mathbb{R}$, i.e., $V$ takes a
state $s \in \mathcal{S}$ and returns a function $V(s) :
\mathbb{R}^{\mathcal{S}} \to \mathbb{R}$ that takes a reward vector $\mathbf{r}
\in \mathbb{R}^{|\mathcal{S}|}$ and returns a value of the state $s$,
$\V(s) \in \mathbb{R}$. The function $V(s)$ computes the values of
all states and returns the value of state $s$.

@@

\begin{lemma} \label{lemma:bound1}
  Let $c : \mathbb{R}^{|\mathcal{S}|} \times \mathbb{R}^m \to (a, b) \subset
  \mathbb{R}$ be an arbitrary bounded function. Then, for $i = 0,
  \dots, d$,
  \[
    \left. \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} \right|_{\lambda_i
      = c(\mathbf{r}, \mathbf{u})}
  \]
  has upper and lower bounds of the form $q(\mathbf{r} \mid \mathbf{u})d(\mathbf{u})$, where
  $d(\mathbf{u}) \in \mathbb{R}_2[\mathbf{u}]$.
\end{lemma}
% TODO: Lemma 1.1 is wrong about the % derivative. Fix (my handwritten notes
% have the correct idea)
\begin{proof}
  Remember that
  \begin{align*}
    \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} &= \frac{1}{2}q(\mathbf{r} \mid \mathbf{u}) (|\bm\Gamma|^{-1} \tr(\mathbf{R} \adj(\bm\Gamma)) \\
                                                                      &- (\mathbf{r} - \mathbf{Su})^\intercal\bm\Gamma^{-1}\mathbf{R}\bm\Gamma^{-1}(\mathbf{r} - \mathbf{Su})).
  \end{align*}
  by Lemma \ref{lemma:derivatives}. We begin by producing constant upper and
  lower bounds for the elements of
  \[
    \left. \frac{\partial \mathbf{K}}{\partial \lambda_i} \right|_{\lambda_i =
      c(\mathbf{r}, \mathbf{u})},
  \]
  where $\mathbf{K}$ is any covariance matrix.

@@

  Combining the bounds with the observation that
  every element of
  $\Kuu^{-1}\mathbf{u}\mathbf{u}^\intercal\Kuu^{-\intercal}$ is in
  $\mathbb{R}_2[\mathbf{u}]$ gives the required result.
\end{proof}

\begin{remark}
  In order to find $\frac{\partial q(\mathbf{u})}{\partial t}$,
  where $t$ is the $i$th element of the vector $\bm\mu$, we can
  find $\frac{\partial q(\mathbf{u})}{\partial \bm\mu}$ and simply take the
  $i$th element. A similar line of reasoning applies to matrices as well. Thus,
  we only need to consider derivatives with respect to $\bm\mu$ and
  $\bm\Sigma$.
\end{remark}

>>>>

% TODO: mention in text
%While the ELBO can now be computed without approximations, we ended up imposing
%two zero-variance restrictions, resulting in a less informative model. Moreover,
%we had to abandon the Bayesian treatment of $\bm\lambda$, making the model more
%vulnerable to overfitting.

@@

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node [rectangle, draw, fill=lightgray] (perturbation) at (2, 0) {\textsc{Perturbation Lemma}};
    \node [rectangle, draw, fill=lightgray] (lebesgue) at (-1, -3) {\textsc{Dominated Convergence Theorem}};
    \node [rectangle, draw] (derivatives) at (-2, 0) {\textsc{Derivatives of PDFs}};
    \node [rectangle, draw] (measurability) at (-3, -2) {\textsc{Measurability}};
    \node [rectangle, draw] (boundedness) at (3, -2) {\textsc{Boundedness}};
    \node [rectangle, draw] (lemma1) at (-2, -1) {\textsc{Lemma \ref{lemma:bound1}}};
    \node [rectangle, draw] (lemma2) at (0, -1) {\textsc{Lemma \ref{lemma:bound2}}};
    \node [rectangle, draw] (lemma3) at (2, -1) {\textsc{Lemma \ref{lemma:bound3}}};
    \node [rectangle, draw] (lemma4) at (3, -3) {\textsc{Lemma \ref{lemma:integral_of_r}}};
    \node [rectangle, draw] (theorem) at (0, -2) {\textsc{Theorem \ref{thm:main}}};

    \draw [arrow] (perturbation) edge (lemma3);
    \draw [arrow] (lebesgue) edge (theorem);
    \draw [arrow] (derivatives) edge (lemma1);
    \draw [arrow] (derivatives) edge (lemma2);
    \draw [arrow] (derivatives) edge (lemma3);
    \draw [arrow] (measurability) edge (theorem);
    \draw [arrow] (boundedness) edge (theorem);
    \draw [arrow] (lemma1) edge (theorem);
    \draw [arrow] (lemma2) edge (theorem);
    \draw [arrow] (lemma3) edge (theorem);
    \draw [arrow] (lemma4) edge (theorem);
  \end{tikzpicture}
  \caption{A graphical representation of dependencies between our theoretical
    results. An arrow from $A$ to $B$ means that $A$ was used to prove $B$.
    Results from the literature are in gray.}
  \label{fig:theorems}
\end{figure}

The typical way to optimise a quantity (the ELBO, in this case) involves
computing its gradient. Unfortunately, some of the terms in $\mathcal{L}$ are
still left as expected values. The goal of this section is to show how Theorem
\ref{thm:lebesgue} can be applied to our model in order to derive the gradient
anyway\footnote{This technique is inspired by black box variational inference
  \cite{DBLP:conf/aistats/RanganathGB14}, but takes a more detailed look at the
  problem and requires significantly more work to prove correctness.}. After
showing that the theorem applies to our situation, given any term expressed as
an expected value $\mathbb{E}[f(\mathbf{u}, \mathbf{r})]$, we can estimate its
gradient as
\[
  \nabla\mathbb{E}[f(\mathbf{u}, \mathbf{r})] = \mathbb{E}[\nabla f(\mathbf{u},
  \mathbf{r})] \approx \frac{1}{S} \sum_{s=1}^S \nabla f(\mathbf{u}_s,
  \mathbf{r}_s),
\]
which can be computed by drawing $S$ Monte Carlo samples $(\mathbf{u}_s,
\mathbf{r}_s) \sim \approximation$.

Our main goal is Theorem \ref{thm:main}, which allows us to move differentiation
inside the integral. In order to prove it, we use a number of intermediate
results. We start by providing a sketch proof of the measurability of MDP value
functions, which is non-obvious due to their non-trivial definition. We then
prove that the value functions are bounded, and establish four helpful lemmas,
three of which provide well-behaved bounds on derivatives of PDFs. See Figure
\ref{fig:theorems} for an overview of how the results fit together.

\begin{remark}
  MDP values are characterised by both a state and a reward function/vector. In
  this section, we think of the value function as $V : \mathcal{S} \to
  \mathbb{R}^{|\mathcal{S}|} \to \mathbb{R}$, i.e., $V$ takes a state $s \in
  \mathcal{S}$ and returns a function $V(s) : \mathbb{R}^{|\mathcal{S}|} \to
  \mathbb{R}$ that takes a reward vector $\mathbf{r} \in
  \mathbb{R}^{|\mathcal{S}|}$ and returns a value of the state $s$, $\V(s) \in
  \mathbb{R}$. Given a reward vector, the function $V(s)$ computes the values of
  all states and returns the value of state $s$.
\end{remark}

@@

\begin{lemma} \label{lemma:bound1}
  Let $c : \mathbb{R}^{|\mathcal{S}|} \times \mathbb{R}^m \to (a, b) \subset
  \mathbb{R}$ be an arbitrary bounded function. Then, for $i = 0,
  \dots, d$,
  \[
    \left. \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} \right|_{\lambda_i
      = c(\mathbf{r}, \mathbf{u})}
  \]
  has upper and lower bounds of the form $q(\mathbf{r} \mid \mathbf{u})d(\mathbf{u})$, where
  $d(\mathbf{u}) \in \mathbb{R}_2[\mathbf{u}]$.
\end{lemma}
\begin{proof}
  Remember that
  \begin{align*}
    \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} &= \frac{1}{2}q(\mathbf{r} \mid \mathbf{u}) (|\bm\Gamma|^{-1} \tr(\mathbf{R} \adj(\bm\Gamma)) \\
                                                                      &- (\mathbf{r} - \mathbf{Su})^\intercal\bm\Gamma^{-1}\mathbf{R}\bm\Gamma^{-1}(\mathbf{r} - \mathbf{Su})).
  \end{align*}
  by Lemma \ref{lemma:derivatives}.

  %We begin by producing constant upper and
  %lower bounds for the elements of
  %\[
  %  \left. \frac{\partial \mathbf{K}}{\partial \lambda_i} \right|_{\lambda_i =
  %    c(\mathbf{r}, \mathbf{u})},
  %\]
  %where $\mathbf{K}$ is any covariance matrix.

@@

  % TODO: Lemma 1.1 is wrong about the % derivative. Fix (my handwritten notes
  % have the correct idea)
  Combining the bounds with the observation that
  every element of
  $\Kuu^{-1}\mathbf{u}\mathbf{u}^\intercal\Kuu^{-\intercal}$ is in
  $\mathbb{R}_2[\mathbf{u}]$ gives the required result.
\end{proof}

\begin{remark}
  In order to find a derivative such as $\frac{\partial q(\mathbf{u})}{\partial
    \mu_i}$, we can find $\frac{\partial q(\mathbf{u})}{\partial \bm\mu}$ and
  simply take the $i$th element. A similar line of reasoning applies to matrices
  as well. Thus, we only need to consider derivatives with respect to $\bm\mu$
  and $\bm\Sigma$.
\end{remark}
